---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Mickey Zhang"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(here); library(tidyverse); library(agricolae); library(ggplot2); library(lubridate)
here()
NTL_LTER_raw = read.csv(
  file = here('Data','Raw','NTL-LTER_Lake_ChemistryPhysics_Raw.csv'),
  stringsAsFactors = TRUE
)
NTL_LTER_raw$sampledate = as.Date(NTL_LTER_raw$sampledate, format = "%m/%d/%y")
  
#2
mytheme = theme_bw()+
  theme(plot.title = element_text(hjust = 0.5),
  plot.subtitle = element_text(hjust = 0.5),
  legend.position = "right"
  )
set_theme(mytheme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer:
\newline
H~0~: Mean July temperature is the same at all depths across lakes.
\newline
H~a~: Mean July temperature differs with depth and/or across lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_LTER = NTL_LTER_raw %>% 
  filter(month(sampledate)== 7) %>%
  select('lakename', 'year4', 'daynum', 'depth','temperature_C') %>%
  drop_na()
  
#5
ggplot(NTL_LTER, aes(x = depth, y = temperature_C))+
  geom_point(size = 0.8)+
  geom_smooth(method = "lm", se = FALSE)+
  scale_y_continuous(expand = c(0,0), limits = c(0, 35)) + 
  scale_x_continuous(expand = c(0,0))+
  labs(
    title = "Temperature by Depth",
    subtitle = "Mean July Temperature",
    x = "Depth (m)",
    y = "Temperature(C)",
  )
  
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The blue regression line suggests a strong negative trend overall. The distribution of points shows some curvature or nonlinearity — temperatures decrease rapidly in the upper few meters, then level off at greater depths.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
LR = lm(data = NTL_LTER, temperature_C ~ depth)
summary(LR)
```

8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Depth explains about 73.9% of the variability (R-squared = 0.7387). The result is based on 9,726 degrees of freedom, showing a large sample size and reliable estimates. The relationship is highly significant because (p < 2.2 × 10^-16^). For every 1-meter increase in depth, the mean July temperature decreases by approximately 1.95°C. 

---
## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
GLM = lm(data = NTL_LTER,
         temperature_C ~ year4 + daynum + depth)
step(GLM)

#10
summary(GLM)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC model selection procedure identified year, day number, and depth as the final set of explanatory variables. This model explains approximately 74.1% of the observed variance in temperature (R-squared = 0.7412). It is slightly improved as R-squared increased from 0.7387 to 0.7412.

## Analysis of Variance
12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
ANOVA = aov(data = NTL_LTER, temperature_C ~ lakename)
summary(ANOVA)
ANOVA2 = lm(data = NTL_LTER, temperature_C ~ lakename)
summary(ANOVA2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Mean lake temperature differs significantly among the nine lakes included in the analysis with F(8,9719)=50, p < 2.2×10^-16^.The extremely small p-value indicates that these differences are highly unlikely to have occurred by chance.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(NTL_LTER, aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(size = 0.8, alpha = 0.5)+
  geom_smooth(method = "lm", se = FALSE)+
  scale_y_continuous(expand = c(0,0), limits = c(0, 35)) + 
  scale_x_continuous(expand = c(0,0))+
  labs(
    title = "Temperature by Depth",
    subtitle = "Mean July Temperature by Lake",
    x = "Depth (m)",
    y = "Temperature(C)",
  )
  
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(ANOVA)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Peter Lake, Paul Lake (p = 0.224), and Ward Lake(p = 0.783) have statistically indistinguishable mean July temperatures (p > 0.05). No single lake stands out as having a mean temperature that is significantly different from all others
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: T-test.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
NTL_LTER = NTL_LTER %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")
t.test(NTL_LTER$temperature_C ~ NTL_LTER$lakename)
```

>Answer: The test does not find a statistically significant difference (p = 0.2649) between the mean temperatures of two lakes. Statistically speaking, we don’t have enough evidence to say they are different, so we treat them as equal. Results from part 16 also suggest that the two lakes have similar mean July temperatures (p = 0.9714459).


